Numerical experiment investigating the influence of drag on trajectory patterns of floating macroalgae

### 2021-03-30 thesis_chapter_3.Rmd
corrects 2....

The study domain falls within the southern Benguela upwelling ecosystem, which is a region of high mesoscale variability and activity. These occur mostly in the form of eddies, filaments and vortices which are generated from the Agulhas retroflection [@veitch2018; @veitch2017], generally flowing northward along the coastline and begin to move offshore near Cape Columbine. The Benguela Jet is a coastal flowing current and is driven by a complex interaction of eddies, upwelling, Eckman transport and geostrophic currents [@stramma1989; @veitch2017; @veitch2018; @hutchings2009]. The stability of the jet varies seasonally with more stable flows in winter and splitting of the jet during summer and spring. In January, the jet begins to split between Cape Point and Cape Columbine [@veitch2018; @blanke2005; @blanke2002] which results into a weak narrow inshore branch which flows along the shelf edge and a strong flowing north-westwards offshore branch. The area is also an area of high upwelling activity driven by strong westerly winds [@hutchings2009]. These winds are most intense during summer (December-March) which favours upwelling conditions in the region, and also act to reinforce the flow of the Benguela Jet; which occurs due to cold upwelling waters contributing to the across-shore density gradient [@ragoasha2019]. In winter the wind direction is reversed which decreases upwelling favorable conditions, but the core of the Benguela jet is maintained by the intrusion of warm Agulhas waters into the Benguela region over the shelf edge [@veitch2018; @hutchings2009; @ragoasha2019].

@@

The model is based on parameterisations that have been used for previous studies investigating iceberg, capsized marine vessels and microplastics. It should be noted that most of the previous studies regarding macroalgae trajectory have investigated rafts and not solitary floating individuals. A kelp raft will vary in size and can range from a few meters across to large dense mats the size of a rugby field. Furthermore, some authors suggest that most macroalgae become entangled through ocean currents and not through the hydrodynamic forces that dislodge them. In addition, large kelp rafts are not a feature around the coast of the Cape Peninsula; rather, solitary kelp are (pers. obs). Therefore, the trajectory of only solitary *E. maxima* individuals will be considered. Numerical calculations of particle trajectory were performed with a Lagrangian method [@delandmeter2019], and be described by the equation

$$X(t+\Delta t) = X(t) + \int_{t}^{t+\Delta t} v(x, \tau)d \tau + \Delta X_b(t)$$
where $X$ is the three-dimensional position of a particle, $v(x,t)$ is the three-dimensional velocity field at the location in the ocean model at particle point in time, $t$ is time, $X_b(t)$ is a change in position due to 'behavior' of the particle at a particular point in time. This can range from swimming in fish to sinking or beaching. In order to investigate the effects of drag components of trajectory of floating macroalgae, this study incorporated drag as the custom behavior.

This study simulated two scenarios; one where the plant is fully submerged and only exposed to hydrodynamic drag; and the other where parts of the plants are partially exposed resulting in a combination of hydrodynamic and air form drag. Drag based on shape-based coefficients were used in the calculations for determining the overall kelp velocity vectors. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, $F_f$ the surface-current flow, and $d_t$ is the time step. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag and surface current-flow are two-dimensional vector quantities and were calculated for each time step for each simulation.

@@

Virtual particles were released with no drag or wind effects from the study site in order to act as a reference for comparison to simulations with varying degrees of hydrodynamic and wind drag, and are referred to hereafter as 'passive particles'. All particles that have any form of drag exposure are hereafter referred to as 'kelp' particles. All simulations have the same release location (Kommetjie) and were run for 30 days with hourly outputs using a fourth order Runge-Kutte advection scheme [@North2009; @VanSebille2018]. Brownian motion was also incorporated in order to account for any stochasticity not resolved by the model. These are standard practices for Lagrangian trajectory studies [@North2009;@delandmeter2019; @VanSebille2018]. To test the different magnitudes that hydrodynamic and wind drag might play in the overall trajectory, simulations were run with varying degrees of drag exposure scenarios. These scenarios were meant to act as proxy for buoyancy, which determines the surface area exposed to hydrodynamic and wind drag. These estimates were expressed as percentages which were used to calculate hydrodynamic and wind drag for applicable simulation. For example, if 85% of the plant is submerged and 15% of above water, then the overall hydrodynamic and wind drag forces would be 85% and 15% of the total drag forces respectively. This approach has been used in past research investigating pelagic Sargassum rafts [@putman2020;@putman2018]. The exposure scenarios used in this study were 100%, 95%, 90%, and 85% for hydrodynamic drag; 0 %, 5%, 10%, and 15% for wind (see table 1). Each scenario consisted of a particle set with a size of 1000 where released from Kommetjie at the start of simulation. The particle sets were advected with a Runge-Kutte fourth order scheme for 30 days and incorporated brownian motion. Since the aim of this investigation was to investigate the relative surface behaviour of virtual kelp particles, only the surface layer was used for simulations. It should also be noted that the Copernicus model used places velocity vectors over some of the land covering the coastal domain. This was not considered a concern, as once again, the aim was to investigate surface drift behavior. The time period chosen to run the simulations was 01-01-2018 un 30-01-2018. This particular time period was chosen as the splitting of the Benguela Jet occurs in January, which also coincides with a period of higher wind intensity, upwelling along the coast and mesoscale activity [@rubio2009; @veitch2018; @ragoasha2019]. A summary of the parameters for each simulation are provided in table 1.

>>>>

The study domain falls within the southern Benguela upwelling ecosystem, which is a region of high mesoscale variability and activity in the form of eddies, filaments and vortices generated from the Agulhas retroflection [@veitch2018; @veitch2017] they generally flow northward along the coastline and begin to move offshore near Cape Columbine. The Benguela Jet is a coastal flowing current and is driven by a complex interaction of eddies, upwelling, Eckman transport and geostrophic currents [@stramma1989; @veitch2017; @veitch2018; @hutchings2009]. The stability of the jet varies seasonally with more stable flows in winter; and in January, the jet begins to split between Cape Point and Cape Columbine [@veitch2018; @blanke2005; @blanke2002] which results in a weak narrow inshore branch that flows along the shelf edge and a strong flowing north-westwards offshore branch. The area is also characterized by high upwelling activity driven by strong westerly winds [@hutchings2009]. These winds are most intense during summer (December-March) which favors upwelling conditions in the region, and also act to reinforce the flow of the Benguela Jet which occurs due to cold upwelling waters contributing to the across-shore density gradient [@ragoasha2019]. In winter the wind direction is reversed which decreases upwelling favorable conditions, but the core of the Benguela jet is maintained by the intrusion of warm Agulhas waters into the Benguela region over the shelf edge [@veitch2018; @hutchings2009; @ragoasha2019].

@@

The model is based on parameterizations that have been used for previous studies investigating iceberg, capsized marine vessels and microplastics. It should be noted that most of the previous studies regarding macroalgae trajectory have investigated rafts and not solitary floating individuals. A kelp raft will vary in size and can range from a few meters across to large dense mats the size of a rugby field. Some authors suggest that most macroalgae become entangled through ocean currents and not through the hydrodynamic forces that dislodge them. Large kelp rafts are not a feature around the coast of the Cape Peninsula; rather, solitary kelp are (pers. obs). Therefore, the trajectory of only solitary *E. maxima* individuals will be considered. We performed numerical calculations of particle trajectory with a Lagrangian method [@delandmeter2019], and be described by the equation

$$X(t+\Delta t) = X(t) + \int_{t}^{t+\Delta t} v(x, \tau)d \tau + \Delta X_b(t)$$
where $X$ is the three-dimensional position of a particle, $v(x,t)$ is the three-dimensional velocity field at the location in the ocean model at particle point in time, $t$ is time $X_b(t)$ is a change in position due to 'behavior' of the particle at a particular point in time. In order to investigate the effects of drag components of trajectory of floating macroalgae, this study incorporated drag as the custom behavior.

This study simulated two scenarios; one where the plant is fully submerged and only exposed to hydrodynamic drag; and the other where parts of the plants are partially exposed resulting in a combination of hydrodynamic and air form drag. Drag based on shape-based coefficients were used in the calculations for determining the overall kelp velocity vectors. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, $F_f$ the surface-current flow, and $d_t$ is the time step. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag, and surface current-flow are two-dimensional vector quantities and were calculated for each time step for each simulation.

@@

Virtual particles were released with no drag or wind effects from the study site in order to act as a reference for comparison to simulations with varying degrees of hydrodynamic and wind drag, and are referred to hereafter as 'passive particles'. All particles that have any form of drag exposure are hereafter referred to as 'kelp particles'. All simulations have the same release location (Kommetjie) and were run for 30 days with hourly outputs using a fourth order Runge-Kutte advection scheme [@North2009; @VanSebille2018]. Brownian motion was also incorporated in order to account for any stochasticity not resolved by the model. These are standard practices for Lagrangian trajectory studies [@North2009;@delandmeter2019; @VanSebille2018]. To test the different magnitudes that hydrodynamic and wind drag might play in the overall trajectory, simulations were run with varying degrees of drag exposure scenarios. These scenarios were meant to act as proxy for buoyancy, which determines the surface area exposed to hydrodynamic and wind drag. These estimates were expressed as percentages which were used to calculate hydrodynamic and wind drag for applicable simulation. For example, if 85% of the plant is submerged and 15% of above water, then the overall hydrodynamic and wind drag forces would be 85% and 15% of the total drag forces respectively. This approach has been used in past research investigating pelagic *Sargassum* rafts [@putman2020;@putman2018]. The exposure scenarios used in this study were 100%, 95%, 90%, and 85% for hydrodynamic drag; 0 %, 5%, 10%, and 15% for wind (see table 1). Each scenario consisted of a particle set with a size of 1000 were released from Kommetjie at the start of simulation. Since the aim of this investigation was to investigate the relative surface behavior of virtual kelp particles, only the surface layer was used for simulations. It should also be noted that the Copernicus model used places velocity vectors over some of the land covering the coastal domain. This was not considered a concern, as once again, the aim was to investigate surface drift behavior. The time period chosen to run the simulations was 01-01-2018 un 30-01-2018. This particular time period was chosen as the splitting of the Benguela Jet occurs in January, which also coincides with a period of higher wind intensity, upwelling along the coast and mesoscale activity [@rubio2009; @veitch2018; @ragoasha2019]. A summary of the parameters for each simulation are provided in table 1.
